---
title: "A Quick and Dirty Guide to S3 Methods in R"
author: "Nicholas Tierney"
date: "29 May 2015"
output:
  word_document: default
  pdf_document: default
  html_document:
    highlight: tango
    keep_md: yes
    theme: readable
---

- Put their contribution in context, in particular discuss related R functions or packages;
- Explain the motivation for their contribution;
- Provide code examples that are reproducible.

S3 methods are a way of writing functions in R that do different things for objects with different classes. 

For example, the function `summary()` used on a dataframe shows descriptive statistics for each variable. For example, the `iris` dataset.

```{r }

summary(iris)

```

But this is different when applied to a linear model

```{r}

lm_iris <- lm(Sepal.Length ~ Sepal.Width, data = iris)

summary(lm_iris)

```

This produces really awesome, different information. It shows what the linear model was, `call`, the `residuals`, `coefficients`, `t-values`, `p-values`, and $R^2$.  It is also **completely** different to the `summary` information we got for the `iris` dataframe.

What's the deal?

R is _hiding_ information from us. There are different functions for `summary`:

- `summary.lm`, for linear models (class = lm)
- `summary.data.frame`, for data frames (class = data.frame)

But you only see the first part - `summary`.

Quick reminder - the class of an object can be found using the command, `class()`, on the object. 

```{r}

x <- c(1, 2, 3, 4, 5)

x

class(x)

```

Here showing that the object `x` is of class `numeric`. `str()` can also provide more information:

```{r}

str(x)

```

Showing that x is `numeric` and then showing its contents.

Back to using S3 methods. When using `summary` on the `iris` data, R actually calls the function `summary.data.frame`. This is because `iris` is of class `data.frame`.

```{r}

class(iris)

```

R is really crafty and hides all that stuff after the `.` on `summary`. Check it out:

```{r}

summary.data.frame(iris)

```

And 

```{r}

summary.lm(lm_iris)

```

Using the wrong class method can give you some output that can be a bit confusing.

```{r}

summary.data.frame(lm_iris)

```

**So why does R hide important stuff after the `.`?**

Because then you'd have to use a different `summary` function for each different class. Imagine how annoying it would be to only use `summary.lm` for linear models, or `summary.data.frame` for data frames. 

## How do I make my own S3 method?

It's quite easy to set up. 

But let's first talk through an example of why you might want to write your own S3 method.

**Case Study of S3 Methods**

I use decision trees. A lot. And something that I do often is calculate the Residual Sums of Squares, $\sum(Y_i - \hat{Y})^2$. Calculating this isn't too difficult, but the way you calculate the answers is different for each decision tree model. So I wrote a function for each decision tree method, but I decided it would be cool to put them inside the one function, so I didn't have to remember if it was "rss_rpart" or "rss_brt" or "rss_random_forest". So I wrote a single function that checks the class of the object being inputted, and then applied different methods dependent upon the class.

This is what I shall now call a "Poor man's S3 method".

```{r}

dt_rss <- function (x){

#CART
if ("rpart" %in% class(x)) {
  
  result <- sum((residuals(x)^2))
  
  return(result)
  
}

# BRT
else if ("gbm" %in% class(x)) {
  
  result <- sum(x$residuals^2)
  
  return(result)

}

# Random Forest

else if ("randomForest" %in% class(x)) {

  temp <- x$finalModel$y - x$finalModel$predicted  
  
  result <- sum(temp^2)  
  
  return(result)

}
  
else warning(paste(class(x), "is not of an rpart, gbm.step, or randomForest object"))
}
##

```


Here's it in action:

```{r}

library(rpart)

fit.rpart <- rpart(Kyphosis ~ Age + Number + Start, data = kyphosis)

dt_rss(fit.rpart)

# not an rpart object

fit.lm <- lm(Sepal.Width ~ Species, data = iris)

dt_rss(fit.lm)

```

We can see that when we specify a class it doesn't know we have to spit out a special message, since it doesn't fit into the classes we have specified.

And sure - it works, but it doesn't make you feel good inside, like good programming does, and isn't as modular - you have this one cluster of ifelses, and that could unnessarily slow down analyses. It is also a little more finnicky to edit. So instead we create functions for each class.

Instead, let's apply a quick and dirty approach to S3 methods.

```{r}

#' RSS
#' \description A function that gives returns the RSS of a decision tree model.
#'
#'
#' Constructor functions

rss <- function(x){
  
  UseMethod("rss", x)
  
}

rss.default <- function(x, ...)
{
  print("You screwed up. I do not know how to handle this object.")

}


#' CART

rss.rpart <- function(x){
  
  sum((residuals(x)^2))
  # cart_ghs_rss <- sum((residuals(rpart.ghs)^2))  
}


#' BRT

rss.gbm <- function(x){
  
  sum(x$residuals^2)
  # brt_ghs_rss <- sum(brt.ghs$residuals^2)  
}


#' Random Forest
#' 

rss.randomForest <- function(x){

  res <- x$finalModel$y - x$finalModel$predicted  
  
  sum(res^2)  
    
#   residual_rf <- rf_ghs_caret$finalModel$y - rf_ghs_caret$finalModel$predicted  
#   rf_ghs_rss <- sum(residual_rf^2)  
}


```

So here we define a method, and then we say what to do of classes .myclass,

```{r}

rss(fit.rpart)

rss(lm.fit)

```


The `.default` is also really cool as it is what tries to work when you have a class that you haven't specified. In this case, this one here is a little insulting.


